---
title: "Annotated Guide to causl Package Code"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Annotated Guide to causl Package Code

## Overview
The `causl` package implements methods for specifying, simulating from, and fitting marginal causal models. It uses a "frugal parameterization" approach where you specify marginal causal quantities of interest and model their dependence structure using copulas. This is particularly useful for causal inference when you want to separate the marginal causal effects from the dependency structure.

---

## Code Block 1: Multivariate Gaussian Approach

```{r multivariate-gaussian, eval=FALSE}
library(causl)

## MULTIVARIATE GAUSSIAN PARAMETERIZATION
# Simpler approach using multivariate normal copula
pars <- list(
  # MARGINAL PARAMETERS (same as before)
  Zc1 = list(beta = c(1), phi = 1),    # Zc1 ~ N(1, 1)
  Zc2 = list(beta = c(1), phi = 1),    # Zc2 ~ N(1, 1)  
  X = list(beta = c(-1,1,1)),           # Treatment model (note: different intercept)
  Y = list(beta = c(1,1), phi = 1),     # Outcome model
  
  # COPULA CORRELATION MATRIX
  cop = list(
    beta = matrix(c(1,0,    # Row 1: correlations for Zc1 (Z3 in matrix notation)
                    1,0,    # Row 2: correlations for Zc2 (Z4 in matrix notation)  
                    0,0),   # Row 3: correlations for Y|do(X)
                  nrow = 2) # 2x3 matrix structure
  )
)

set.seed(1)  # For reproducible results

# FAMILY SPECIFICATION (simpler structure)
fams <- list(c(1,1),  # Both covariates are Gaussian
            5,        # Treatment is Bernoulli
            1,        # Outcome is Gaussian  
            1)        # Copula is Gaussian

# GENERATE DATA
data_samples_mvg <- causalSamp(
  10000,  # Sample size
  formulas = list(
    list(Zc1~1, Zc2~1),  # Both covariates are marginal (intercept only)
    X ~ Zc1 + Zc2,       # Treatment depends on covariates
    Y ~ X,               # Outcome depends on treatment
    ~ X                  # Copula depends on treatment
  ), 
  family = fams, 
  pars = pars
)
```

**Key Points:**

- Uses `causalSamp()` instead of `rfrugalParam()`
- Simpler correlation matrix approach for dependencies
- Larger sample size (10,000) for more stable estimates

---

## Code Block 2: Vine Copula Approach

```{r vine-copula, eval=FALSE}
## VINE COPULA SPECIFICATION
# More detailed pair copula approach with explicit vine structure
fams <- list(
  rep(1,2),  # Both covariates (Zc1, Zc2) are Gaussian
  5,         # Treatment X is Bernoulli  
  1,         # Outcome Y is Gaussian
  1          # Copula family is Gaussian
)

# FORMULA STRUCTURE
forms <- list(
  list(Zc1 ~ 1, Zc2 ~ 1),  # Both covariates are marginal
  X ~ Zc1 + Zc2,           # Treatment model
  Y ~ X,                   # Outcome model  
  ~ 1                      # Copula intercept
)

# DETAILED PARAMETER SPECIFICATION
pars <- list(
  # MARGINAL DISTRIBUTIONS
  Zc1 = list(beta = 1, phi = 1),       # Zc1 ~ N(1, 1)
  Zc2 = list(beta = 1, phi = 1),       # Zc2 ~ N(1, 1)
  
  # TREATMENT MODEL
  X = list(beta = c(-2,1,1)),           # logit(P(X=1)) = -2 + Zc1 + Zc2
  
  # OUTCOME MODEL  
  Y = list(beta = c(1,1), phi = 1),     # Y = 1 + X + error, error ~ N(0,1)
                                        # This is the CAUSAL effect: Y|do(X)
  
  # VINE COPULA SPECIFICATION
  cop = list(
    Y = list(
      Zc1 = list(beta = c(1)),          # Dependency: Y and Zc1 
      Zc2 = list(beta = c(0))           # No dependency: Y and Zc2
    )
  )
)

# SIMULATION WITH VINE COPULA
data_samples_vinecop <- rfrugalParam(
  n = 1e3,           # 1000 samples
  formulas = forms,   # Model structure
  family = fams,      # Distribution families
  pars = pars         # Parameters
)
```

**Key Points:**

- This version uses vine copulas for more flexible dependency modeling
- The `cop` parameter specifies which variables are dependent through the copula
- `beta = c(1)` means strong positive dependence, `beta = c(0)` means independence
- The link function for copula parameters is `2*expit(beta) - 1`

---

## Code Block 3: Pair Copula Approach (Original Version)

```{r pair-copula-v1, eval=FALSE}
library(causl)

## FAMILY SPECIFICATION
# Defines the distributional families for each component of the model
fams <- list(
  rep(1,1),           # Covariate Zc1: Gaussian (family code 1)
  5,                  # Treatment X: Bernoulli (family code 5) 
  c(1,1),            # Two components, both Gaussian
  list(rep(1,1), rep(1,2))  # Nested list structure for copula families
)

## PARAMETER SPECIFICATION
# Contains all parameters needed for the model components
pars <- list(
  # MARGINAL PARAMETERS FOR COVARIATES
  Zc1 = list(beta = 1, phi = 1),      # Zc1 ~ N(1, 1) - mean=1, sd=1
  Zc2 = list(beta = 1, phi = 1),      # Zc2 ~ N(1, 1) - mean=1, sd=1
  
  # TREATMENT MODEL PARAMETERS  
  X = list(beta = c(-2,1,1)),          # Logistic: logit(P(X=1)) = -2 + 1*Zc1 + 1*Zc2
  
  # OUTCOME MODEL PARAMETERS
  Y = list(beta = c(1,1), phi = 1),    # Linear: Y = 1 + 1*X + error, error ~ N(0,1)
  
  # COPULA PARAMETERS (dependency structure)
  cop = list(
    Zc2 = list(
      Zc1 = list(beta = c(1))          # Dependency between Zc1 and Zc2
    ),
    Y = list(
      Zc1 = list(beta = c(1)),         # Dependency between Y and Zc1
      Zc2 = list(beta = c(0))          # No dependency between Y and Zc2 (beta=0)
    )
  )
)

## FORMULA SPECIFICATION
# Defines the structural relationships in the causal model
forms <- list(
  list(Zc1 ~ 1),      # Zc1 has only intercept (marginal distribution)
  X ~ Zc1 + Zc2,      # Treatment depends on both covariates
  list(Zc2 ~ 1,       # Zc2 has only intercept
       Y ~ X),        # Outcome depends on treatment
  ~ 1                 # Copula specification (intercept only)
)

# SIMULATION
# Generate 1000 samples from the specified causal model
rfrugalParam(n = 1000, formulas = forms, family = fams, pars = pars)
```

**Key Points:**

- This uses a more complex pair copula approach to model dependencies
- The copula parameters control how variables are correlated beyond their marginal distributions
- More nested structure in both `fams` and `cop` parameters
- Family codes: 1=Gaussian, 5=Bernoulli, 3=Gamma, etc.

---

## Understanding the Causal Framework

### Model Structure
All three approaches simulate from the same basic causal model:
```
Zc1, Zc2 (covariates) → X (treatment) → Y (outcome)
Zc1, Zc2 → Y (direct effects, controlled by copula)
```

### Key Concepts

1. **Frugal Parameterization**: Separates marginal causal effects from dependency structure
2. **Copula**: Models the dependence between variables while preserving marginal distributions
3. **Causal Effect**: The `Y ~ X` relationship represents `E[Y|do(X)]`, the causal effect of treatment
4. **Confounding**: Controlled through the copula parameters that link covariates to outcomes

### Function Differences
- `rfrugalParam()`: More flexible, handles complex copula structures
- `causalSamp()`: Simpler interface, good for standard multivariate normal dependencies

### Family Codes
- 1: Gaussian/Normal
- 3: Gamma/Exponential  
- 5: Bernoulli/Binomial
- Others available for different distributions

This framework is particularly useful for sensitivity analysis in causal inference, where you want to test how robust your causal conclusions are to different assumptions about unmeasured confounding.